Accelerating structure–function mapping using the ViVa webtool to mine natural variation

### 2018-11-15 manuscript/02-results.Rmd
update viva use outline

## An introduction to ViVa 
**Morgan will provide screenshots**

@@

Perhaps the most critical domain is the degron of the Aux/IAA transcriptional repressor proteins. Mutations which prevent this domain from being recognized and marked for proteasomal degradation by SCF^TIR1/AFB^ are dominant and often result in obvious phenotypes [@berleth_auxin_2004; @yang_iaa1_2004]. As expected this critical functional region is highly conserved at the amino acid level \@ref(fig:A-IAA-alignment). Similarly, the EAR motif which facilitates Aux/IAA interaction with TPL/TPR transcriptional repressors, are also conserved.

@@

(ref:A-IAA-alignment) **Canonical IAAs have conserved Degron and EAR motifs** Protein sequences were aligned [@wright_decipher_2015] and variants were mapped to this alignment and colored according to the predicted functional effect of the most deleterious variant at that position, with light colors being less deleterious and darker colors being more deleterious. Red indicates nonsynonymous variants. Color scale is explained in {#methods}. EAR motif is highlighted by a black box including alignment positions 70-74. Degron domain is highlighted by a gray box including alignment positions 194-201.
```{r A-IAA-alignment, fig.cap="(ref:A-IAA-alignment)"}
A_IAAs <- subset(gene_info, tair_symbol %in% IAA_data[which(IAA_data$Class == "A"), "name"])
A_IAAs <- alignCDS(A_IAAs$transcript_ID)
# make an alignment data frame
A_IAAs <- makeAlnDF(A_IAAs[[2]])
A_IAAs <- left_join(A_IAAs, dplyr::select(gene_info, "tair_locus",
                                       "tair_symbol", "transcript_ID"),
                        by = c("transcript_ID" = "transcript_ID"))
A_IAAs$tair_symbol %<>% as.factor()
A_IAAs$tair_symbol %<>% fct_reorder(
    {as.integer(str_remove(A_IAAs$tair_symbol, 
                           pattern = "IAA"))}, 
    .desc = TRUE)
VCF <- ldply(.data = VCF, .fun = subset, gt_GT != "0|0")
coding_VCF <- getCodingDiv(VCF)
A_IAAs <- addSNPsToAlnDF(aln_df = A_IAAs,SNPs = coding_VCF, effect_order = SNPeff_order)
A_IAAs$gap <- A_IAAs$seq_pos == "-"

@@

The Aux/IAA proteins bind to ARF transcription factors allowing the Aux/IAA's to effect transcriptional repression at specific loci. Thus in the ARF transcription factor proteins we expect the DNA-binding function to be conserved. Indead, the DNA-binding domains of the ARFs had very few, low-diversity missense polymorphisms, as did the PB1 domain which facilitates ARF-Aux/IAA interaction \@ref(fig:A-ARF-Alignment).

@@

(ref:A-ARF-Alignment) **Canonical ARFs have conserved DNA binding and PB1 domains** Protein sequences were aligned [@wright_decipher_2015] and variants were mapped to this alignment and colored according to the predicted functional effect of the most deleterious variant at that position, with light colors being less deleterious and darker colors being more deleterious. Red indicates nonsynonymous variants. Color scale is explained in {#methods}. 
```{r A-ARF-Alignment, fig.cap="(ref:A-ARF-Alignment)"}
A_ARFs <- subset(gene_info, tair_symbol %in% c("ARF5", "ARF6", "ARF7", "ARF8", "ARF19"))
A_ARFs <- alignCDS(A_ARFs$transcript_ID)
# make an alignment data frame
A_ARFs <- makeAlnDF(A_ARFs[[2]])

>>>>

## An introduction to ViVa

The current version of ViVa is specifically focused on visualizes natural variation in the coding sequences of genes. Future versions will expand upon this to examine noncoding sequences; however, indels are much more common in noncoding sequences, which limits the certainty of this resequencing data.

**Morgan will provide screenshots**
### Data input: Select the desired genes

### SNPstats: Summary of gene information, structure, and diversity

### Diversity plot: Visualize allelic diversity across the coding sequence

### Accession mapping

### Alignment

### Gene tree

### SNP browser

@@

Perhaps the most critical domain is the degron of the Aux/IAA transcriptional repressor proteins. Mutations which prevent this domain from being recognized and marked for proteasomal degradation by SCF^TIR1/AFB^ are dominant and often result in obvious phenotypes [@berleth_auxin_2004; @yang_iaa1_2004]. As expected this critical functional region is highly conserved at the amino acid level (\@ref(fig:A-IAA-alignment)). Similarly, the EAR motif which facilitates Aux/IAA interaction with TPL/TPR transcriptional repressors, are also conserved.

@@

(ref:A-IAA-alignment) **Canonical IAAs have conserved Degron and EAR motifs.** Protein sequences were aligned [@wright_decipher_2015] and variants were mapped to this alignment and colored according to the predicted functional effect of the most deleterious variant at that position, with light colors being less deleterious and darker colors being more deleterious. Red indicates missense variants. Color scale is explained in [Methods](#methods). EAR motif is highlighted by a black box including alignment positions 70-74. Degron domain is highlighted by a gray box including alignment positions 194-201.
```{r A-IAA-alignment, fig.cap="(ref:A-IAA-alignment)"}
A_IAAs <- subset(gene_info, tair_symbol %in% IAA_data[which(IAA_data$Class == "A"), "name"])
A_IAAs <- alignCDS(A_IAAs$transcript_ID)
# make an alignment data frame
A_IAAs <- makeAlnDF(A_IAAs[[2]])
A_IAAs <- left_join(A_IAAs, dplyr::select(gene_info, "tair_locus",
                                       "tair_symbol", "transcript_ID"),
                        by = c("transcript_ID" = "transcript_ID"))
A_IAAs$tair_symbol %<>% as.factor()
A_IAAs$tair_symbol %<>% fct_reorder(
    {as.integer(str_remove(A_IAAs$tair_symbol, 
                           pattern = "IAA"))}, 
    .desc = TRUE)
VCF <- ldply(.data = VCF, .fun = subset, gt_GT != "0|0")
coding_VCF <- getCodingDiv(VCF)
A_IAAs <- addSNPsToAlnDF(aln_df = A_IAAs,SNPs = coding_VCF, effect_order = SNPeff_order)
A_IAAs$gap <- A_IAAs$seq_pos == "-"

@@

The Aux/IAA proteins bind to ARF transcription factors allowing the Aux/IAA's to effect transcriptional repression at specific loci. Thus in the ARF transcription factor proteins we expect the DNA-binding function to be conserved. Indead, the DNA-binding domains of the ARFs had very few, low-diversity missense polymorphisms, as did the PB1 domain which facilitates ARF-Aux/IAA interaction (Figure \@ref(fig:A-ARF-Alignment)).

@@

(ref:A-ARF-Alignment) **Canonical ARFs have conserved DNA binding and PB1 domains.** Protein sequences were aligned [@wright_decipher_2015] and variants were mapped to this alignment and colored according to the predicted functional effect of the most deleterious variant at that position, with light colors being less deleterious and darker colors being more deleterious. Red indicates missense variants. Color scale is explained in [Methods](#methods).
```{r A-ARF-Alignment, fig.cap="(ref:A-ARF-Alignment)"}
A_ARFs <- subset(gene_info, tair_symbol %in% c("ARF5", "ARF6", "ARF7", "ARF8", "ARF19"))
A_ARFs <- alignCDS(A_ARFs$transcript_ID)
# make an alignment data frame
A_ARFs <- makeAlnDF(A_ARFs[[2]])

### 2018-11-15 manuscript/03-methods.Rmd
update viva use outline

where $N_{ik}$ is the nucleotide (A, T, C or G) at position k on the ith sequence of the population.
$L$ is the length of the sequence (the same for all sequences in the population, 
assumes only SNPs exist, no indels). n is the total number of sequences in the population.

@@

The function `Nucleotide_diversity` in the r1001genomes package calculates $\pi_k$ for each position in the gene or region that contains a variant. note, $\pi_k$ is equal to 0 at all locations without variants.
This is also what is displayed in the Diversity Plot tab of the web app.

## Detailed $\pi_k$ calculaiton simplification.

@@

\begin{equation}
\displaystyle \pi_k = \sum_{ij} \frac{n_i}{n}\frac{n_j}{n} \pi_{ij} = 
\frac{1}{n^2}\sum_{ij} n_i n_j \pi_{ij}\\
\pi_{ij(k)} = \left\{\begin{array}{l}
1 \quad \textrm{if} \quad i \neq j \\
0 \quad \textrm{if} \quad i = j
\end{array}\right.
(\#eq:nei-simp1)
\end{equation}

>>>>

where $N_{ik}$ is the nucleotide (A, T, C or G) at position k on the *i*th sequence of the population.
$L$ is the length of the sequence (the same for all sequences in the population, 
assumes only SNPs exist, no indels). n is the total number of sequences in the population.

@@

The function `Nucleotide_diversity` in the r1001genomes package calculates $\pi_k$ for each position in the gene or region that contains a variant. Note, $\pi_k$ is equal to 0 at all locations without variants.
This is also what is displayed in the Diversity Plot tab of the web app.

## Detailed $\pi_k$ calculation simplification.

@@

\begin{equation}
\displaystyle \pi_k = \sum_{ij} \frac{n_i}{n}\frac{n_j}{n} \pi_{ij} = \frac{1}{n^2}\sum_{ij} n_i n_j \pi_{ij}
\qquad
\pi_{ij(k)} = \left\{\begin{array}{l}
1 \quad \textrm{if} \quad i \neq j \\
0 \quad \textrm{if} \quad i = j
\end{array}\right.
(\#eq:nei-simp1)
\end{equation}

### 2018-11-15 manuscript/index.Rmd
update viva use outline

abstract: Thousands of sequenced genomes are now publically available and represent a large swath of plant evolution; yet, much of this data remains inaccessible to researchers without significant bioinformatics experience. Here, we present a webtool called ViVa (Visualization of Variation) which aims to empower any researcher to take advantage of the amazing genetic resource collected in the Arabidopsis 1001 Genomes Project. ViVa facilates data mining on the gene, gene family or gene network level. To test the utility and accessibility of ViVa, we assembled a team with a range of expertise with biology and bioinformatics to analyze the natural variation in the well-studied nuclear auxin signaling pathway. Our analysis has provided further confirmation of existing knowledge and has also helped generate new hypotheses regarding this well studied pathway. These results highlight how natural variation can be used to generate and test hypotheses about less studied gene families and networks, especially when paired with biochemical and genetic characterization. ViVa is also readily extensible to newly emerging databases of genetic variation, such as the 3,000 Rice Genomes Project.

>>>>

abstract: Thousands of sequenced genomes are now publically available and represent a large swath of plant evolution; yet, much of this data remains inaccessible to researchers without significant bioinformatics experience. Here, we present a webtool called ViVa (Visualizing Variation) which aims to empower any researcher to take advantage of the amazing genetic resource collected in the Arabidopsis 1001 Genomes Project. ViVa facilates data mining on the gene, gene family or gene network level. To test the utility and accessibility of ViVa, we assembled a team with a range of expertise with biology and bioinformatics to analyze the natural variation in the well-studied nuclear auxin signaling pathway. Our analysis has provided further confirmation of existing knowledge and has also helped generate new hypotheses regarding this well studied pathway. These results highlight how natural variation can be used to generate and test hypotheses about less studied gene families and networks, especially when paired with biochemical and genetic characterization. ViVa is also readily extensible to newly emerging databases of genetic variation, such as the 3,000 Rice Genomes Project.
